# (PART\*) Part A. Exploring Data {-}

# Visualizing Data {#dataviz}

Part A of these Notes is designed to ease your transition into working effectively with data, so that you can better understand it. We'll start by visualizing some data from the US National Health and Nutrition Examination Survey, or NHANES. We'll display R code as we go, but we'll return to all of the key coding ideas involved later in the Notes.

## The NHANES data: Collecting a Sample

To begin, we'll gather a random sample of 1,000 subjects participating in NHANES, and then identify several variables of interest about those subjects^[For more on the NHANES data available in the NHANES package, type ?NHANES in the Console in R Studio.]. Some of the motivation for this example came from a Figure in @BaumerKaplanHorton.

```{r create_nh_dat1}
# library(NHANES) # already loaded NHANES package/library of functions, data

set.seed(431001) 
# use set.seed to ensure that we all get the same random sample 
# of 1,000 NHANES subjects in our nh_data collection

nh_dat1 <- sample_n(NHANES, size = 1000) %>%
    select(ID, Gender, Age, Height) 

nh_dat1
```

We have `r nrow(nh_dat1)` rows (observations) and `r ncol(nh_dat1)` columns (variables) that describe the subjects listed in the rows. 

## Age and Height

Suppose we want to visualize the relationship of Height and Age in our 1,000 NHANES observations. The best choice is likely to be a scatterplot.

```{r nh_dat1_heightbyage1-fig}
ggplot(data = nh_dat1, aes(x = Age, y = Height)) +
    geom_point()
```

We note several interesting results here.

1. As a warning, R tells us that it has "Removed `r 1000 - sum(!is.na(nh_dat1$Age) & !is.na(nh_dat1$Height))` rows containing missing values (geom_point)." Only `r sum(!is.na(nh_dat1$Age) & !is.na(nh_dat1$Height))` subjects plotted here, because the remaining `r 1000 - sum(!is.na(nh_dat1$Age) & !is.na(nh_dat1$Height))` people have missing (NA) values for either Height, Age or both. 
2. Unsurprisingly, the measured Heights of subjects grow from Age `r min(nh_dat1$Age)` to Age 20 or so, and we see that a typical Height increases rapidly across these Ages. The middle of the distribution at later Ages is pretty consistent at at a Height somewhere between 150 and 175. The units aren't specified, but we expect they must be centimeters. The Ages are clearly reported in Years.
3. No Age is reported over 80, and it appears that there is a large cluster of Ages at 80. This may be due to a requirement that Ages 80 and above be reported at 80 so as to help mask the identity of those individuals.^[If you visit the NHANES help file with ?NHANES, you will see that subjects 80 years or older were indeed recorded as 80.]

As in this case, we're going to build most of our visualizations using tools from the `ggplot2` package, which is part of the `tidyverse` series of packages. You'll see similar coding structures throughout this Chapter, most of which are covered as well in Chapter 3 of @R4DS.


## Subset of Subjects with Known Age and Height

Before we move on, let's manipulate the data set a bit, to focus on only those subjects who have complete data on both Age and Height. This will help us avoid that warning message.

```{r create_nh_dat2_attempt1}
nh_dat2 <- nh_dat1 %>%
    filter(complete.cases(Age, Height)) 

summary(nh_dat2)
```

Note that the units and explanations for these variables are contained in the NHANES help file, available via typing `?NHANES` in the Console of R Studio, or by typing `NHANES` into the Search bar in R Studio's Help window.

### The Distinction between `Gender` and `Sex`

The `Gender` variable here is a mistake. These data refer to the biological status of these subjects, which is their `Sex`, and not the social construct of `Gender` which can be quite different. In our effort to avoid further confusion, we'll rename the variable `Gender` to instead more accurately describe what is actually measured here. 

To do this, we can use this approach...

```{r create_nhdat2_corrected}
nh_dat2 <- nh_dat1 %>%
    rename(Sex = Gender) %>%
    filter(complete.cases(Age, Height)) 

summary(nh_dat2)
```

That's better. How many observations do we have now? We could use `dim` to find out the number of rows and columns in this new data set.

```{r}
dim(nh_dat2)
```

Or, we could simply list the data set and read off the result.

```{r}
nh_dat2
```

## Age-Height and Sex?

Let's add Sex to the plot using color, and also adjust the y axis label to incorporate the units of measurement. 

```{r nh_dat2_heightbyageandsex1-fig}
ggplot(data = nh_dat2, aes(x = Age, y = Height, color = Sex)) +
    geom_point() +
    labs(title = "Height-Age Relationship in NHANES sample", 
         y = "Height in cm.")
```

### Can we show the Female and Male relationships in separate panels?

Sure.

```{r nh_dat2_heightbyageandsex2-fig}
ggplot(data = nh_dat2, aes(x = Age, y = Height, color = Sex)) +
    geom_point() + 
    labs(title = "Height-Age Relationship in NHANES sample", 
         y = "Height in cm.") +
    facet_wrap(~ Sex)
```

### Can we add a smooth curve to show the relationship in each plot?

Yep, and let's change the theme of the graph to remove the gray background, too.

```{r nh_dat2_heightbyageandsex3-fig}
ggplot(data = nh_dat2, aes(x = Age, y = Height, color = Sex)) +
    geom_point() + 
    geom_smooth(method = "loess") +
    labs(title = "Height-Age Relationship in NHANES sample", 
         y = "Height in cm.") +
    theme_bw() +
    facet_wrap(~ Sex)
```


### What if we want to assume straight line relationships?

We could look at a linear model in the plot. Does this make sense here?

```{r nh_dat2_heightbyageandsex4-fig}
ggplot(data = nh_dat2, aes(x = Age, y = Height, color = Sex)) +
    geom_point() + 
    geom_smooth(method = "lm") +
    labs(title = "Height-Age Relationship in NHANES sample", 
         y = "Height in cm.") +
    theme_bw() +
    facet_wrap(~ Sex)
```

## A New Subset: Ages 21-79

Suppose we wanted to look only at those observations (subjects) whose Age is at least 21 and at most 79. Suppose also that we want to look at some of the additional variables available in NHANES. To start, we'll do the following:

1. Set a seed for the random sampling from the original NHANES data, so we get the same original sample of 1000 people we started with earlier.
2. Select 1000 people from the NHANES data.
3. Filter the sample to only those people whose age is more than 20 and less than 80 years.
4. Select the variables we will be using in the remainder of this chapter, which will be:
    - `Age` as we've seen before, in years.
    - `Height` as we've seen before, in centimeters.
    - `Gender` which we'll rename as `Sex` again.
    - `Pulse` = 60 second pulse rate (in beats per minute).
    - `BPSysAve` = Systolic Blood Pressure, in mm Hg (and we'll rename this `SBP`).
    - `SleepTrouble` = Yes means the subject has told a health professional that they had trouble sleeping.
    - `PhysActive` = Yes means the subject does moderate or vigorous-intensity sports, fitness or recreational activity.
    - `MaritalStatus` = one of Married, Widowed, Divorced, Separated, NeverMarried or LivePartner (living with partner.)
    - `HealthGen` = self-reported rating of general health, one of Excellent, Vgood (Very Good), Good, Fair or Poor.
5. Rename `Gender` as `Sex`, to more accurately describe what is being measured.
6. Omit subjects with any missingness on *any* of the variables we've selected. 

Can you see how the code below accomplishes these tasks?

```{r identify nh_data subjects ages 21-79}
set.seed(431001) # again, this will ensure the same sample

nh_data_2179 <- sample_n(NHANES, size = 1000) %>%
    filter(Age > 20 & Age < 80) %>%
    select(ID, Gender, Age, Height, 
           Pulse, BPSysAve, SleepTrouble, PhysActive,
           MaritalStatus, HealthGen) %>%
    rename(Sex = Gender, SBP = BPSysAve) %>%
    na.omit

nh_data_2179
```

## Distribution of Heights

What is the distribution of height in this new sample?

```{r nh_data_2179_heighthistogram-fig}
ggplot(data = nh_data_2179, aes(x = Height)) + 
    geom_histogram() 
```

We can do several things to clean this up. 

1. We'll change the color of the lines for each bar of the histogram.
2. We'll change the fill inside each bar to make them stand out a bit more.
3. We'll add a title and relabel the horizontal (x) axis to include the units of measurement.
4. We'll avoid the warning by selecting a number of bins (we'll use 25 here) into which we'll group the heights before drawing the histogram.

```{r nh_data_2179_heighthistogram2-fig}
ggplot(data = nh_data_2179, aes(x = Height)) + 
    geom_histogram(bins = 25, col = "yellow", fill = "blue") + 
    labs(title = "Height of NHANES subjects ages 21-79",
         x = "Height in cm.")
```

### Changing a Histogram's Fill and Color

The CWRU color guide (https://case.edu/umc/our-brand/visual-guidelines/) lists the HTML color schemes for CWRU blue and CWRU gray. Let's match that color scheme.

```{r nh_data_2179_histogramwithCWRUscheme-fig}
cwru.blue <- '#0a304e'
cwru.gray <- '#626262'

ggplot(data = nh_data_2179, aes(x = Height)) + 
    geom_histogram(binwidth = 2, col = cwru.gray, fill = cwru.blue) + 
    labs(title = "Height of NHANES subjects ages 21-79",
         x = "Height in cm.") +
    theme_bw()
```

Note the other changes to the graph above.

1. We changed the theme to replace the gray background.
2. We changed the bins for the histogram, to gather observations into groups of 2 cm. each.


## Height and Sex

```{r nh_data_2179_heightbysex1-fig}
ggplot(data = nh_data_2179, aes(x = Sex, y = Height, color = Sex)) + 
    geom_point() + 
    labs(title = "Height by Sex for NHANES subjects ages 21-79",
         y = "Height in cm.")
```

This plot isn't so useful. We can improve things a little by jittering the points horizontally, so that the overlap is reduced.

```{r nh_data_2179_heightbysex2-fig}
ggplot(data = nh_data_2179, aes(x = Sex, y = Height, color = Sex)) + 
    geom_jitter(width = 0.2) + 
    labs(title = "Height by Sex (jittered) for NHANES subjects ages 21-79",
         y = "Height in cm.")
```

Perhaps it might be better to summarise the distribution in a different way. We might consider a boxplot of the data.

### A Boxplot of Height by Sex

```{r nh_data_2179_heightbysexbox-fig}
ggplot(data = nh_data_2179, aes(x = Sex, y = Height, fill = Sex)) + 
    geom_boxplot() + 
    labs(title = "Boxplot of Height by Sex for NHANES subjects ages 21-79",
         y = "Height in cm.")
```

Or perhaps we'd like to see a pair of histograms?

### Histograms of Height by Sex

```{r nh_data_2179_heightbysexhist-fig}
ggplot(data = nh_data_2179, aes(x = Height, fill = Sex)) + 
    geom_histogram(color = "white", bins = 20) + 
    labs(title = "Histogram of Height by Sex for NHANES subjects ages 21-79",
         x = "Height in cm.") + 
    facet_wrap(~ Sex)
```

Can we redraw these histograms so that they are a little more comparable, and to get rid of the unnecessary legend?

```{r nh_data_2179_heightbysexhist2-fig}
ggplot(data = nh_data_2179, aes(x = Height, fill = Sex)) + 
    geom_histogram(color = "white", bins = 20) + 
    labs(title = "Histogram of Height by Sex for NHANES subjects ages 21-79 (Revised)",
         x = "Height in cm.") + 
    guides(fill = FALSE) +
    facet_grid(Sex ~ .)
```

## Looking at Pulse Rate

Let's look at a different outcome, the *pulse rate* for our subjects.

Here's a histogram, again with CWRU colors, for the pulse rates in our sample.

```{r nh_data_2179_pulse_histbin1-fig}
ggplot(data = nh_data_2179, aes(x = Pulse)) + 
    geom_histogram(binwidth = 1, fill = cwru.blue, col = cwru.gray) + 
    labs(title = "Histogram of Pulse Rate: NHANES subjects ages 21-79",
         x = "Pulse Rate (beats per minute)")
```

Suppose we instead bin up groups of 5 beats per minute together as we plot the Pulse rates. 

```{r nh_data_2179_pulse_histbin5-fig}
ggplot(data = nh_data_2179, aes(x = Pulse)) + 
    geom_histogram(binwidth = 5, fill = cwru.blue, col = cwru.gray) + 
    labs(title = "Histogram of Pulse Rate: NHANES subjects ages 21-79",
         x = "Pulse Rate (beats per minute)")
```

Which is the more useful representation will depend a lot on what questions you're trying to answer.

### Pulse Rate and Physical Activity

We can also split up our data into groups based on whether the subjects are physically active. Let's try a boxplot.

```{r nh_data_2179_pulse_by_activity_box-fig}
ggplot(data = nh_data_2179, aes(y = Pulse, x = PhysActive, fill = PhysActive)) + 
    geom_boxplot() + 
    labs(title = "Pulse Rate by Physical Activity Status for NHANES ages 21-79")
```

As an accompanying numerical summary, we might ask how many people fall into each of these `PhysActive` categories, and what is their "average" `Pulse` rate.

```{r}
nh_data_2179 %>%
    group_by(PhysActive) %>%
    summarise(count = n(), mean(Pulse), median(Pulse)) %>%
    knitr::kable(digits = 2) 
```

The `knitr::kable(digits = 2)` piece of this command tells R Markdown to generate a table with some attractive formatting, and rounding any decimals to two figures.

### Pulse by Sleeping Trouble

```{r nh_data_2179_pulse_by_sleep_histbin1-fig}
ggplot(data = nh_data_2179, aes(x = Pulse, fill = SleepTrouble)) + 
    geom_histogram(color = "white", bins = 20) + 
    labs(title = "Histogram of Pulse Rate by Sleep Trouble for NHANES subjects ages 21-79",
         x = "Pulse Rate (beats per minute)") + 
    guides(fill = FALSE) +
    facet_grid(SleepTrouble ~ ., labeller = "label_both")
```


How many people fall into each of these `SleepTrouble` categories, and what is their "average" Pulse rate?

```{r, include = TRUE}
nh_data_2179 %>%
    group_by(SleepTrouble) %>%
    summarise(count = n(), mean(Pulse), median(Pulse)) %>%
    knitr::kable(digits = 2) 
```


### Pulse and HealthGen

We can compare the distribution of Pulse rate across groups by the subject's self-reported overall health (`HealthGen`), as well.

```{r nh_data_2179_pulsebyhealthgen1-fig}
ggplot(data = nh_data_2179, aes(x = HealthGen, y = Pulse, fill = HealthGen)) + 
    geom_boxplot() +
    labs(title = "Pulse by Self-Reported Overall Health for NHANES ages 21-79",
         x = "Pulse Rate") + 
    guides(fill = FALSE) 
```    

How many people fall into each of these `HealthGen` categories, and what is their "average" Pulse rate?

```{r}
nh_data_2179 %>%
    group_by(HealthGen) %>%
    summarise(count = n(), mean(Pulse), median(Pulse)) %>%
    knitr::kable(digits = 2) 
```

### Pulse Rate and Systolic Blood Pressure

```{r nh_data_2179_pulsevssbp-fig}
ggplot(data = nh_data_2179, aes(x = SBP, y = Pulse)) +
    geom_point() +
    geom_smooth(method = "loess") +
    labs(title = "SBP vs. Pulse rate for NHANES subjects, ages 21-79")
```


### Sleep Troubls vs. No Sleep Trouble?

Could we see whether subjects who have described `SleepTrouble` show different SBP-pulse rate patterns than the subjects who haven't? 

- Let's try doing this by changing the shape *and* the color of the points based on `SleepTrouble`.

```{r nh_data_2179_sbpvspulsewithsleep-fig}
ggplot(data = nh_data_2179, 
       aes(x = SBP, y = Pulse, 
           color = SleepTrouble, shape = SleepTrouble)) +
    geom_point() +
    geom_smooth(method = "loess") +
    labs(title = "SBP vs. Pulse rate for NHANES subjects, ages 21-79")
```

This plot might be easier to interpret if we faceted by `SleepTrouble`, as well.

```{r nh_data_2179_sbpvspulsewithsleepfacets-fig}
ggplot(data = nh_data_2179, 
       aes(x = SBP, y = Pulse, 
           color = SleepTrouble, shape = SleepTrouble)) +
    geom_point() +
    geom_smooth(method = "loess") +
    labs(title = "SBP vs. Pulse rate for NHANES subjects, ages 21-79") +
    facet_wrap(~ SleepTrouble, labeller = "label_both")
```

## General Health Status

Here's a Table of the General Health Status results. Again, this is a self-reported rating of each subject's health on a five point scale (Excellent, Very Good, Good, Fair, Poor.) 

```{r table_HealthGen1}
nh_data_2179 %>%
    select(HealthGen) %>%
    table()
```

The HealthGen data are categorical, which means that summarizing them with averages isn't as appealing as looking at percentages, proportions and rates.

Another, somewhat simpler way to get a table of this sort of information uses the `tabyl` function from the `janitor` package in R.


```{r tabyl_HealthGen2}
# tabyl is part of the janitor package
# already loaded: library(janitor)

nh_data_2179 %>%
    tabyl(HealthGen) 
```

I don't actually like the title of `percent` here, as it's really a proportion, but that can be adjusted, and we can add a total.

```{r tabyl_HealthGen3}
nh_data_2179 %>%
    tabyl(HealthGen) %>%
    adorn_totals() %>%
    adorn_pct_formatting()
```

When working with an unordered categorical variable, like `MaritalStatus`, the same approach can work.

```{r tabyl_MaritalStatus1}
nh_data_2179 %>%
    tabyl(MaritalStatus) %>%
    adorn_totals() %>%
    adorn_pct_formatting()
```

### Bar Chart for Categorical Data

Usually, a **bar chart** is the best choice for a graphing a variable made up of categories.

```{r HealthGengraph1-fig}
ggplot(data = nh_data_2179, aes(x = HealthGen)) + 
    geom_bar()
```

There are lots of things we can do to make this plot fancier.

```{r HealthGengraph2-fig}
ggplot(data = nh_data_2179, aes(x = HealthGen, fill = HealthGen)) + 
    geom_bar() + 
    guides(fill = FALSE) +
    labs(x = "Self-Reported Health Status",
         y = "Number of NHANES subjects",
         title = "Self-Reported Health Status in NHANES subjects ages 21-79")
```

Or, we can really go crazy...

```{r HealthGengraph3-fig}
nh_data_2179 %>%
    count(HealthGen) %>%
    ungroup() %>%
    mutate(pct = round(prop.table(n) * 100, 1)) %>%
    ggplot(aes(x = HealthGen, y = pct, fill = HealthGen)) + 
    geom_bar(stat = "identity", position = "dodge") +
    scale_fill_viridis_d() +
    guides(fill = FALSE) +
    geom_text(aes(y = pct + 1,    # nudge above top of bar
                  label = paste0(pct, '%')),  # prettify
              position = position_dodge(width = .9), 
              size = 4) +
    labs(x = "Self-Reported Health Status",
         y = "Percentage of NHANES subjects",
         title = "Self-Reported Health Status in NHANES subjects ages 21-79") +
    theme_bw()
```


### Working with Tables

We can add both row and column marginal totals, and compare subjects by Sex, as follows...

```{r tabHealthSex2nokable-tab}
nh_data_2179 %>%
    tabyl(Sex, HealthGen) %>%
    adorn_totals(c("row", "col")) 
```

If we like, we can make this look a little more polished with the `knitr::kable` function...

```{r tabHealthSex2nokable-tab2}
nh_data_2179 %>%
    tabyl(Sex, HealthGen) %>%
    adorn_totals(c("row", "col")) %>%
    knitr::kable()
```

Or, we can get a complete cross-tabulation, including (in this case) the percentages of people within each Sex that fall in each HealthGen category (percentages within each row) like this.

```{r tabHealthSex2nokable-tab3}
nh_data_2179 %>%
    tabyl(Sex, HealthGen) %>%
    adorn_totals("row") %>%
    adorn_percentages("row") %>%
    adorn_pct_formatting() %>%
    adorn_ns() %>%
    knitr::kable()
```

And, if we wanted the column percentages, to determine which sex had the higher rate of each HealthGen status level, we can get that by changing the adorn_percentages to describe results at the column level:

```{r tabHealthSex2nokable-tab4}
nh_data_2179 %>%
    tabyl(Sex, HealthGen) %>%
    adorn_totals("col") %>%
    adorn_percentages("col") %>%
    adorn_pct_formatting() %>%
    adorn_ns() %>%
    knitr::kable()
```

### SBP by General Health Status

Let's consider now the relationship between self-reported overall health and systolic blood pressure.

```{r nh_data_2179_sbpbyhealth-fig}
ggplot(data = nh_data_2179, aes(x = HealthGen, y = SBP, fill = HealthGen)) + 
    geom_boxplot() + 
    labs(title = "SBP by Health Status, Overall Health for NHANES ages 21-79",
         y = "Systolic Blood Pressure", x = "Self-Reported Overall Health") + 
    guides(fill = FALSE) 
```

We can see that not too many people self-identify with the "Poor" health category.

```{r}
nh_data_2179 %>%
    group_by(HealthGen) %>%
    summarise(count = n(), mean(SBP), median(SBP)) %>%
    knitr::kable() 
```

### SBP by Physical Activity and General Health Status

We'll build a panel of boxplots to try to understand the relationships between Systolic Blood Pressure, General Health Status and Physical Activity. Note the use of `coord_flip` to rotate the graph 90 degrees, and the use of `labeller` within `facet_wrap` to include both the name of the (Physical Activity) variable and its value.

```{r nh_data_2179_bmibyhealthbysex1-fig}
ggplot(data = nh_data_2179, aes(x = HealthGen, y = SBP, fill = HealthGen)) + 
    geom_boxplot() + 
    labs(title = "SBP by Health Status, Overall Health for NHANES ages 21-79",
         y = "Body-mass index", x = "Self-Reported Overall Health") + 
    guides(fill = FALSE) +
    facet_wrap(~ PhysActive, labeller = "label_both") + 
    coord_flip()
```

### SBP by Sleep Trouble and General Health Status

Here's a plot of faceted histograms, which might be used to address similar questions related to the relationship between Overall Health, Systolic Blood Pressure and Sex.

```{r nh_data_2179_bmibyhealthbysex2-fig}
ggplot(data = nh_data_2179, aes(x = SBP, fill = Sex)) + 
    geom_histogram(color = "white", bins = 20) + 
    labs(title = "SBP by Sex, Overall Health for NHANES ages 21-79",
         x = "Body-mass index") + 
    guides(fill = FALSE) +
    facet_grid(HealthGen ~ Sex)
```


## Conclusions

This is just a small piece of the toolbox for visualizations that we'll create in this class. Many additional tools are on the way, but the main idea won't change. Using the `ggplot2` package, we can accomplish several critical tasks in creating a visualization, including:

- Identifying (and labeling) the axes and titles
- Identifying a type of `geom` to use, like a point, bar or histogram
- Changing fill, color, shape, size to facilitate comparisons
- Building "small multiples" of plots with faceting

Good data visualizations make it easy to see the data, and `ggplot2`'s tools make it relatively difficult to make a really bad graph.
